Episodic memory: mental time travel or a quantum `memory wave' function?

### 2019-01-04 main.tex
working through revised outline

Although the specific content of each moment of the episode is largely unique, participants' subjective experiences of the episode entail weaving together those temporally separated moments into a cohesive narrative.  This is reflected in how people verbally recount the episode later.  If we examine the temporal autocorrelation matrices of participants' recalls of the television episode (characterized using the same topic model used to capture the content of the episode; Figs.~\ref{fig:corrmats}B, C), we see a different structure than that of the episode itself.  Although individual participants' recalls exhibit periods of content stability punctuated by rapid change (block diagonal structure of the matrices in Fig.~\ref{fig:corrmats}C), the off diagonal correlations between temporally distant events are substantially stronger in participants' recalls than in the original episode.  This indicates that the way participants \textit{recount} temporally separated events imposes additional similarity on those events, beyond what was present in the original events themselves.  Is this simply a matter of imprecision in recall, either with respect to participants' recalls themselves, or with respect to the quality with which they are characterized via semantic feature vectors?  Or might there be a functional role to that additional imposed similarity structure?

To examine which specific events are described (during recall) as more similar than what their original experiences (as characterized by semantic feature vectors) would suggest, it can be helpful to examine individual recalls of particular events throughout the episode.  \textbf{JRM STOPPED HERE...}

%where i was going with this: is that structure meaningful?  one can ask which specific events are described similarly (reinstatement fig).  argument: yes, the structure is meaningful-- example: the way people describe the particular event (sherlock surmising a new murder has happened between bantering with Watson) is similar to how participants describe the other murders and other sherlock/watson banterings

\begin{figure}[tp] \centering \includegraphics[width=\textwidth]{figs/reinstatement} \caption{\textbf{Content drift in a television episode and participants' recalls of that episode.  A.  Autocorrelations in the episode's content and in the content of participants' recalls.}  The black line denotes the average correlations between the topic vectors from pairs of moments in the episode as a function of their temporal separation (times are relative to the full length of the episode).  Each colored line denotes an analogous autocorrelation function, but for each individual participant's recalls of the episode (times are relative to the full length of each participant's recalls).  The error ribbons denote 95\% confidence intervals, taken across all moments in the episode (or in participants' recalls of the episode).  \textbf{B. Recall density functions for one recalled moment.} Each curve reflects the correlations between the topic vector at each (relative) moment of the episode (or participants' recalls of the episode), and the topic vector of an event beginning approximately $\frac{1}{3}$ through the episode (relative time: 0.328).  The gray curve denotes these correlations for the topic vectors derived from annotations of the episode itself. \textbf{JRM STOPPED HERE...}}
  % colors: individual participants
  % black: average across participants
  % gray: autocorrelation with topic vector at timepoint
  % circles: local maxima in the across-participants average (i.e., movie events that share structure with the given timepoint that is distinct from other temporally proximal events)
\label{fig:reinstatement}
\end{figure}

\section*{BRAINSTORMING OUTLINE OF THE REST OF THE PAPER}
\begin{enumerate}
  \item Show video correlation matrix + recall correlation matrix.  Key idea: off-diagonal mass is introduced (primarily) by how people \textit{remember} the movie, not due to the contents of the movie itself (evidence: correlation matrices).  Further, this is not simply a matter of imprecision in how people recall, and/or our approach's ability to characterize recalled events (evidence: reinstatement figure showing which scenes are recalled similarly).  This sort of integration is how we ascribe meaning to events that unfold over disjointed extended timescales.
  \item In theory this could also happen during \textit{any} sort of experience.  For example, when people study random word lists, they often re-order their recalls so that semantically related words are closer in the recall sequence than they were in the presented sequence.  This can be captured to an extent by semantic clustering measures.  However, it's difficult to examine this phenomenon in more depth because word lists necessarily lack the richness of more naturalistic stimuli.  There are no ``deeper ties'' between study events (i.e., individual words), so the sorts of integration operations that allow us to make sense of real-world experiences aren't given the same freedom of expression in list-learning experiments.  Evidence: presentation/recall correlation matrices during random free recall don't show the same sort of off-diagonal increase.  Claim: mental time travel and quantum wave reinstatement look the same for random word list remembering, but very different for real-world remembering.
  \item The specific ties between unfolding events forms a scaffolding for understanding how those events are related.  This can be visualized geometrically by projecting word embedding representations of the events onto lower dimensional spaces (although see Flatland Fallacy paper, with HyperTools paper as a potential counterpoint).  This reveals another key property of how we remember: although there are substantial individual differences in the specific words people use to describe a shared experience, and how much detail they get into, the overall ``shapes'' of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for how we communicate memories to other people: we need to convey the ``gist'' (shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas (cite Baldassano et al) unfold in similar ways (e.g. have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).
    \item These ideas also have implications for other forms of learning (e.g. spatial learning) and neural decoding.  When we attempt to characterize thoughts (e.g. training a classifier to distinguish between states), we tend to train decoders to estimate the ``one'' state that the brain is in.  However, perhaps we should instead consider that our brains are contemplating, or holding in our thoughts, many possible states simultaneously.  Therefore we should consider the full distribution of states (rather than only considering its maximum) as a primary metric of characterizing mental function.  (Probably people have proposed this-- e.g. representing the posterior and using it to compute...need citations.)
    \end{enumerate}

>>>>

Although the specific content of each moment of the episode is largely unique, participants' subjective experiences of the episode entail weaving together those temporally separated moments into a cohesive narrative.  This is reflected in how people verbally recount the episode later.  If we examine the temporal autocorrelation matrices of participants' recalls of the television episode (characterized using the same topic model used to capture the content of the episode; Figs.~\ref{fig:corrmats}B, C), we see a different structure than that of the episode itself.  Although individual participants' recalls exhibit periods of content stability punctuated by rapid change (block diagonal structure of the matrices in Fig.~\ref{fig:corrmats}C), the off diagonal correlations between temporally distant events are substantially stronger in participants' recalls than in the original episode.  Another way of characterizing this phenomenon is to look at the autocorrelation functions (describing the correlation between different moments' topic vectors as a function of their temporal distance) for the episode and participants' recalls (Fig.~\ref{fig:reinstatement}A).  Whereas the autocorrelations for the original episode rapidly fall to 0, autocorrelations in the content of participants' recalls appear to asymptote around 0.2.  These analyses indicate that the way participants \textit{recount} temporally separated events imposes additional similarity on those events, beyond what was present in the original events themselves.  One can then ask: is this simply a matter of imprecision in recall, either with respect to participants' recalls themselves, or with respect to the quality with which they are characterized via semantic feature vectors?  Or might there be a functional role to that additional imposed similarity structure?

To examine which specific events are described (during recall) as more similar than what their original experiences (as characterized by semantic feature vectors) would suggest, it can be helpful to examine individual recalls of particular events throughout the episode.  Figure~\ref{fig:reinstatement}B displays, for a representative example recalled moment, which other moments from the episode were described in a similar way by the participants when they recounted the episode later (black and colored lines), and which other moments in the episode were described in a similar way in moment-by-moment (non-recall) human annotations (gray line).  The example event explored in the figure comprises a scene where Sherlock Holmes and John Watson (main characters) are bantering, and then are interrupted when Sherlock realizes that another victim has been murdered.  The language used by human annotators describing the moment-by-moment contents of the scene used a similar mix of themes only for moments in the episode that were temporally proximal to the event in question (gray line).  However, the way participants \textit{described} what happened when they recounted the episode later revealed a much different pattern.  Specifically, their language mirrored that used when they were recalling other murders, and other moments when Holmes and Watson were bantering (Fig.~\ref{fig:reinstatment}C).  This suggests that these events were associated in participants' memories, perhaps contributing to the understanding that those events were linked in the episode's narrative.

\begin{figure}[tp] \centering \includegraphics[width=\textwidth]{figs/reinstatement} \caption{\textbf{Content drift in a television episode and participants' recalls of that episode.  A.  Autocorrelations in the episode's content and in the content of participants' recalls.}  The black line denotes the average correlations between the topic vectors from pairs of moments in the episode as a function of their temporal separation (times are relative to the full length of the episode).  Each colored line denotes an analogous autocorrelation function, but for each individual participant's recalls of the episode (times are relative to the full length of each participant's recalls).  The error ribbons denote 95\% confidence intervals, taken across all moments in the episode (or in participants' recalls of the episode).  \textbf{B. Recall density functions for one recalled moment.} Each curve reflects the correlations between the topic vector at each (relative) moment of the episode (or participants' recalls of the episode), and the topic vector of an event beginning approximately $\frac{1}{3}$ through the episode (relative time: 0.328).  The gray curve denotes these correlations for the topic vectors derived from annotations of the episode itself. The black line denotes the average correlations across all participants' recalls.  Local maxima are marked with colored dots.  Local maxima closer than 2.5\% of the total recall duration to a higher local maximum were excluded, as were maxima with peaks with correlation coefficients lower than 0.1 (arbitrarily chosen; these ).   The colored lines denote the correlations for each individual participant's recalls.  \textbf{C. Local maxima in the average recall density function.}  Representative screen captures from the scenes at each local maxima in Panel B are displayed (dot colors match those in Panel B).  The text in each screen capture displays a brief summary of what happened in each scene.}
  % colors: individual participants
  % black: average across participants
  % gray: autocorrelation with topic vector at timepoint
  % circles: local maxima in the across-participants average (i.e., movie events that share structure with the given timepoint that is distinct from other temporally proximal events)
\label{fig:reinstatement}
\end{figure}

\section*{BRAINSTORMING OUTLINE OF THE REST OF THE PAPER}
\begin{enumerate}
  \item \textit{Show video correlation matrix + recall correlation matrix.  Key idea: off-diagonal mass is introduced (primarily) by how people \textit{remember} the movie, not due to the contents of the movie itself (evidence: correlation matrices).  Further, this is not simply a matter of imprecision in how people recall, and/or our approach's ability to characterize recalled events (evidence: reinstatement figure showing which scenes are recalled similarly).  This sort of integration is how we ascribe meaning to events that unfold over disjointed extended timescales.}
  \item In theory this could also happen during \textit{any} sort of experience.  For example, when people study random word lists, they often re-order their recalls so that semantically related words are closer in the recall sequence than they were in the presented sequence.  This can be captured to an extent by semantic clustering measures.  However, it's difficult to examine this phenomenon in more depth because word lists necessarily lack the richness of more naturalistic stimuli.  There are no ``deeper ties'' between study events (i.e., individual words), so the sorts of integration operations that allow us to make sense of real-world experiences aren't given the same freedom of expression in list-learning experiments.  Evidence: presentation/recall correlation matrices during random free recall don't show the same sort of off-diagonal increase.  Claim: mental time travel and quantum wave reinstatement look the same for random word list remembering, but very different for real-world remembering.
  \item The specific ties between unfolding events forms a scaffolding for understanding how those events are related.  This can be visualized geometrically by projecting word embedding representations of the events onto lower dimensional spaces (although see Flatland Fallacy paper, with HyperTools paper as a potential counterpoint).  This reveals another key property of how we remember: although there are substantial individual differences in the specific words people use to describe a shared experience, and how much detail they get into, the overall ``shapes'' of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for how we communicate memories to other people: we need to convey the ``gist'' (shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas (cite Baldassano et al) unfold in similar ways (e.g. have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).
    \item These ideas also have implications for other forms of learning (e.g. spatial learning) and neural decoding.  When we attempt to characterize thoughts (e.g. training a classifier to distinguish between states), we tend to train decoders to estimate the ``one'' state that the brain is in.  However, perhaps we should instead consider that our brains are contemplating, or holding in our thoughts, many possible states simultaneously.  Therefore we should consider the full distribution of states (rather than only considering its maximum) as a primary metric of characterizing mental function.  (Probably people have proposed this-- e.g. representing the posterior and using it to compute...need citations.)
    \end{enumerate}
